A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-10-25 jacobian_calc_CMAME.tex
small fix
spelling and typo cleanup

The way an AD system works is by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, in a way no different from how a normal computer program or human would evaluate composition functions.  AD departs in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their partial derivatives may be computed and stored in a sequence. The AD system then multiples the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence.  In the literature, the AD scheme described here is called \emph{forward automatic-differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, particularly \cite{ref-on-AD} which is foundational. Appendix~\ref{sec:appendixB} includes an example that walks the reader through the process a computer uses to compute derivates via AD.  Some things to note about AD are that no approximation of derivatives is being made because the analytical forms of the partials of the elementary math functions are defined alongside them. The accuracy of AD is then limited by the precision of the AD system's definition of the elementary math functions and their partial derivatives .

@@

In solid and structural computational mechanics the tangent-stiffness matrix is a linearization operator that describes the stiffness of a system in response to small displacements imposed upon the current configuration of the system.  Mathematically speaking, the tangent-stiffness represents the gradient of a high-dimensional energy surface the ``points'' in the minimum direction. While we've adopted the terminology of solid mechanics, it's important to note that these operators appear in other physical settings and are known by other names, e.g. a \emph{transmisability matrix} in the context of a Poisson problem or, more generally, a \emph{Jacobian matrix} in mathematical optimization.

@@

In detail, a tangent-stiffness matrix can be described as a collection of the first-order partial derivatives of a vector valued function with respect to each independent vector component of the function for a given vector.   A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$,
can be expressed in indicial notation as 
%
\begin{equation} 
  K_{ij} = \left. \frac{\partial F_i}{\partial X_j}\right|_{\vec{X}_0}
\end{equation}
%
Where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function, $X_j$ is the $j^{\mbox{th}}$ component of the vector argument to $F$, and a particular value of the vector is chosen to linearize about $\vec{X}_0$. One then evaluates the expression for each combination $i, j$ corresponding to an (row, column) location in the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be estimated using the any of the complex-step, AD, or common finite-difference techniques for functions $F:R^1 \rightarrow R^1$, since taking partial derivatives entails holding all but a single independent vector component of the function constant, and each element of the function can be evaluated independently of the others. \\

@@

When the using the forward-difference technique the formula for calculating a tangent-stiffness
matrix is:

@@

Where $K_{ij}$ is the index notation representation of an element of the tangent-stiffness matrix at
row $i$, column $j$. The expression says that the partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the quotient of the partial difference of two terms
and a denomenator. The first of the 'two terms' is $F_i$ evaluated at $x$ where
the $j$'th component of $x$ has been perturbed by an very small amount in the direction of the
positive $x_j$ axis. The second of the 'two terms' is $F_I$ evaluated at an unperturbed $x$. The
denominator is the magnitude of perturbation called "probe distance" \cite{ref-Adaggio}

@@

Where our model $F$ is treated as function of complex variables and $x$ is treated as a vector of
complex values. The expressions says that the real valued partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the imaginary component of $F_i$ evaluated at $x$
where $x$ is perturbed by a very small amount in the direciton of the imaginary axis corresponding
to the $j$'th component of $x$ all divided by the magnitude of the perturbation. There is no partial
notation because a partial difference is not taken.

@@

Complex step was implemented in \emph{Perdigm} as follows. First a way of persistently allocating
memory needed to store imaginary values for nodal position and dilatation was determined. Persistent
memory allocation is a very basic way to improve performance in general for a computer program. The
way memory was allocated for similar vector and scalar variables used by the prior existing forward
difference method was copied and new 'field spec' variables were added into the 'elastic' material
model to store imaginary position coordinates and imaginary dilatation. Because of \emph{Peridigm}'s
dependence on the \emph{Epetra} package of the \emph{Trilinos} library, 'field spec' variables could
not be simply be declared as having a complex datatype. Program methods that were used in the
course of applying the forward-difference method were copied, renamed and re-written to follow the
CTSE formula, which involved changing the datatype of intermediate variables which were locally
scoped and dynamically casting persistent variables, like the 'field spec' variables, to complex
datatypes where necessary. Additional modifications were for the purpose of collecting experimental
data and are not part of complex-step itself. The mathematical difference between the original
forward-difference and complex-step code was covered in \ref{tsmat}. The source code for
implementing complex-step in \emph{Perdigim} can be found on the corresponding author's website.
Because \emph{Peridigm} was written by computer scientists rather than mechanical engineers
primarily, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic
and distributed data structures were used in the basic code and by necessity in the modified code.
To see complex-step in a more narrow context, it is advised that readers also take a look at the
examples of complex-step and the other methods provided on the corresponding author's website.

@@

The methods were compared by running two sets of test problems where each method would solve a
problem concurrently. Each test problem simulated a $4 by .25 by .25$ meter block of metal
undergoing tension along the axis parallel to the long dimension of the block. The test problems
were set up in \emph{Perdigm} as qausistatic equillibrium problems, where the displacement boundary
conditions used to apply tension were applied gradually, in 'load steps', and an equilibrium
solution for each load step was achieved before applying the next load step. The purpose of applying
the load in gradual steps is a widely used technique to ensure the Newton-Raphson method is always
guessing near the actual solution. The material properties of the model were set at $8.0E-9 metric
tonne/ mm^3$ density, $1.515E4 MPa$ bulk modulus, and $7.813E4 MPa$ shear modulus. The peridynamic
horizon used in a test problem was three times the node spacing for the corrisponding model mesh.
This value for horizon was chosen out of experience to insure that every node could have at least
one nonlocal connection along the diagonal of the node grid.

@@

What was recorded during the test runs as experimental data is the subject the following
subsubsection. It is organized to explain "what", "why" and "how" in that order.

@@

Because it makes no sense to compare tangent-stiffness matrices from different problems, load steps
that started with difference guesses, or from different iterations, it was necessary to solve one
load step and conclude each iteration within that load step by updating the displacement guess only
from the results of the AD method and to subsequently feed all four methods the same updated guess
the following iteration. This decision precluded a comparison of convergence rate, since if the
methods were allowed to solve a problem at their individual pace, differences in accuracy would
produce differences in guess updates and therefor the number and nature of iterations perfomed.
Different guesses from iterations started with different previous guesses could not be data for a
valid comparison of tangent-stiffness accuracy between methods.  However a separate convergence rate
comparison study could be done given some modifications to the materials for the study described in
this paper, and of course, the resources to perform that study.   Additionally, having each of the
methods physically operate in the same process, single or one of associated parallel, allowed the
comparison of tangent-stiffness matrices as loaded from random access memory rather than from the
hard-disk.  Loading from RAM gives the benefit of vastly greater speed and the simpler programming
compared to some other solution speculatively involving dynamic file management on files which for
one of the components of the study would be on the order of $1XE10$ bytes. However, the price of
running the methods together in the same process and avoiding the use of the hard-disk was that at
least two tangent-stiffness matrices had to be stored in RAM during simulation, which meant that
special high-memory compute nodes were needed for the $1E6$ peridynamic element problem series.

@@

Mesh generation for the test problems run in the study was done using the software package
\emph{Cubit} developed at Sandia National Labs \cite{ref-Cubit}. For the test series in which peridynamic node density was
varied, journal scripts compatible with \emph{Cubit} were used. The rule used to select the
arangement of nodes within the borders designated by simulated material block's dimensions was to
manually choose a node spacing, fit the maximum whole number of nodes in a regular grid within the
block, and then check whether or not this number of nodes matched the number of nodes desired for a
particular test. The same process was used for the multicore series. An example journal script can
be found on the corresponding author's website. For future work, it may be beneficial to automate
and randomize this process for statistic reasons and for time efficiency.

@@

\subsubsection{Methods: Reduction of Data} The output produced by \emph{Perdigm} including the
additional accuracy and speed measurements it was modified to provide was redirected from the
console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).
Directories and filenames were chosen so that data would be easy to sort by individual test run.
Extraneous information was manually deleted from copies versions of the output text files so that
they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts
averaged accuracy or speed data for a test corresponding to a particular peridynamic node density
over all iterations for that run, and then plotted this averaged data as if it were a function of
peridynamic node density as it varied from test to test in the series.  Similarly, for the multicore
tests, data was averaged and plotted in much the same way, but as a funciton of number of cores used
to solve the problem rather than as a function peridynamic node density which was held constant.
These data reduction and plotting scripts can be found on the corresponding autor's website.

@@

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in \ref{tcs}. The averaged
results of these accuracy measurements for the single core tests can be seen in
\ref{fig:serial_accuracy}. Averaged Frobenius norm of the element-wise difference between a given
Jacobian and the Jacobian produced by the automatic-differentiation method is plotted against number
of nonzero Jacobian matrix elements. Complex-step was revealed as the most accurate of the methods
besides automatic-differentiation, complex-step resulting in several orders of magnitude less
difference that resulted from using the central-difference method and the most inaccurate
forward-difference method. This accuracy ranking order was mantained for each test in the series. It
should be pointed out compared to the L2 norms of any of the Jacobian matrices by themselves, the
magnitude of these differences for any of the methods is relatively small. These values appear in
the experimental data which is available on the corresponding author's website. The effect of these
accuracy differences on convergence rate directly was not determined.

@@

\subsection{Final Thoughts on Complex-Step} Complex-step was shown to be highly accurate in
calculating tangent-stiffness matrices by the standards of this study, but as it is implemented in
\emph{Peridigm} may not be any more accurate than AD, and it is almost certainly slower for a naive
implementation. When looking to the future of \emph{Peridigm}, Complex-step holds the advantage of
relying on byte-copiable datatypes, while AD requires a complicated serialization and
deserialization in order to function as byte-copiable. This is fine for copying in between MPI
processes, but deserialization on an accelerator or GPU is very difficult, as this would require
compiling a version of the AD library being used that were compatible with the accelerator or GPU
being used. What this means is that if \emph{Peridigm} or any simulation software were to be
reformulated to take advantage of accelerators or GPUs, attractive given their performance and power
efficiency advantages over traditional CPUs, the flexibilty of complex-step owed to its simple
implementation may make it more viable fo rthat application.

>>>>

The way an AD system works is by first evaluating the innermost function of the composition, then presenting that function's output as input to the next level function until all levels are complete, in a way no different from how a normal computer program or human would evaluate composition functions.  AD departs in that as each nested function is evaluated, the function's partial derivative with respect to the designated variables of the given input is also calculated.  This is possible because the elementary math functions are hard-coded into the AD source-code along with their analytical derivatives, and linked by special instructions, so that when the elementary math functions are called upon for computation, their partial derivatives may be computed and stored in a sequence. The AD system then multiplies the final sequence of partial derivatives together to produce the exact equivalent to taking a partial derivative of the corresponding composition function with respect to a designated variable at a particular value. It is obvious, but bears mentioning, that the AD system could simply store one value for partial derivatives, modifying it as appropriate for every function evaluation rather than keeping a sequence.  In the literature, the AD scheme described here is called \emph{forward automatic-differentiation}. Only forward AD will be covered here since it is the implementation used in \emph{Sacado} and therefore this study; however, the reader is referred to the introduction section of \cite{ref-AD-methods} and its citations for further information on AD, particularly \cite{ref-on-AD} which is foundational. Appendix~\ref{sec:appendixB} includes an example that walks the reader through the process a computer uses to compute derivates via AD.  Some things to note about AD are that no approximation of derivatives is being made because the analytical forms of the partials of the elementary math functions are defined alongside them. The accuracy of AD is then limited by the precision of the AD system's definition of the elementary math functions and their partial derivatives .

@@

In solid and structural computational mechanics the tangent-stiffness matrix is a linearization operator that describes the stiffness of a system in response to small displacements imposed upon the current configuration of the system.  Mathematically speaking, the tangent-stiffness represents the gradient of a high-dimensional energy surface that ``points'' in the minimum direction. While we've adopted the terminology of solid mechanics, it's important to note that these operators appear in other physical settings and are known by other names, e.g. a \emph{transmisability matrix} in the context of a Poisson problem or, more generally, a \emph{Jacobian matrix} in mathematical optimization.

@@

In detail, a tangent-stiffness matrix can be described as a collection of the first-order partial derivatives of a vector valued function with respect to each independent vector component of the function for a given vector.   A general mathematical formula for a tangent-stiffness matrix, $K_{ij}$,
can be expressed in indicial notation as 
%
\begin{equation} 
  K_{ij} = \left. \frac{\partial F_i}{\partial X_j}\right|_{\vec{X}_0}
\end{equation}
%
Where $F_i$ is the $i^{\mbox{th}}$ component of the vector valued function, $X_j$ is the $j^{\mbox{th}}$ component of the vector argument to $F$, and a particular value of the vector is chosen to linearize about $\vec{X}_0$. One then evaluates the expression for each combination $i, j$ corresponding to a (row, column) location in the tangent-stiffness matrix. By inspection, the elements of a tangent-stiffness matrix can be estimated using any of the complex-step, AD, or common finite-difference techniques for functions $F:R^1 \rightarrow R^1$, since taking partial derivatives entails holding all but a single independent vector component of the function constant, and each element of the function can be evaluated independently of the others. \\

@@

When using the forward-difference technique the formula for calculating a tangent-stiffness
matrix is:

@@

Where $K_{ij}$ is the index notation representation of an element of the tangent-stiffness matrix at
row $i$, column $j$. The expression says that the partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the quotient of the partial difference of two terms
and a denomenator. The first of the 'two terms' is $F_i$ evaluated at $x$ where
the $j$'th component of $x$ has been perturbed by an very small amount in the direction of the
positive $x_j$ axis. The second of the 'two terms' is $F_I$ evaluated at an unperturbed $x$. The
denominator is the magnitude of perturbation called "probe distance" \cite{ref-Adaggio}.

@@

Where our model $F$ is treated as function of complex variables and $x$ is treated as a vector of
complex values. The expression says that the real valued partial derivative of the $i$'th component of $F$
w.r.t the $j$'th component of $x$ is equal to the imaginary component of $F_i$ evaluated at $x$
where $x$ is perturbed by a very small amount in the direciton of the imaginary axis corresponding
to the $j$'th component of $x$ all divided by the magnitude of the perturbation. There is no partial
notation because a partial difference is not taken.

@@

Complex step was implemented in \emph{Perdigm} as follows. First a way of
persistently allocating memory needed to store imaginary values for nodal
position and dilatation was determined. Persistent memory allocation is a very
basic way to improve performance in general for a computer program. The way
memory was allocated for similar vector and scalar variables used by the prior
existing forward difference method was copied and new 'field spec' variables
were added into the 'elastic' material model to store imaginary position
coordinates and imaginary dilatation. Because of \emph{Peridigm}'s dependence
on the \emph{Epetra} package of the \emph{Trilinos} library, 'field spec'
variables could not be simply be declared as having a complex datatype. Program
methods that were used in the course of applying the forward-difference method
were copied, renamed and re-written to follow the CTSE formula, which involved
changing the datatype of intermediate variables which were locally scoped and
dynamically casting persistent variables, like the 'field spec' variables, to
complex datatypes where necessary. Additional modifications were for the
purpose of collecting experimental data and are not part of complex-step
itself. The source code for implementing complex-step in \emph{Perdigim} can be
found on the corresponding author's website.  Because \emph{Peridigm} was
written by computer scientists rather than mechanical engineers primarily, many
advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic and distributed data structures were used in the basic code and by
necessity in the modified code.  To see complex-step in a more narrow context,
it is advised that readers also take a look at the examples of complex-step and
the other methods provided on the corresponding author's website.

@@

The methods were compared by running two sets of test problems where each method would solve a
problem concurrently. Each test problem simulated a $4$ by $.25$ by $.25$ meter block of metal
undergoing tension along the axis parallel to the long dimension of the block. The test problems
were set up in \emph{Perdigm} as quasistatic equillibrium problems, where the displacement boundary
conditions used to apply tension were applied gradually, in 'load steps', and an equilibrium
solution for each load step was achieved before applying the next load step. The purpose of applying
the load in gradual steps is a widely used technique to ensure the Newton-Raphson method is always
guessing near the actual solution. The material properties of the model were set at $8.0E-9$ $ metric
tonne/ mm^3$ density, $1.515E4$ $MPa$ bulk modulus, and $7.813E4$ $MPa$ shear modulus. The peridynamic
horizon used in a test problem was three times the node spacing for the corrisponding model mesh.
This value for horizon was chosen out of experience to insure that every node could have at least
one nonlocal connection along the diagonal of the node grid.

@@

What was recorded during the test runs as experimental data is the subject the following
subsubsection.

@@

Because it makes no sense to compare tangent-stiffness matrices from different problems, load steps
that started with difference guesses, or from different iterations, it was necessary to solve one
load step and conclude each iteration within that load step by updating the displacement guess only
from the results of the AD method and to subsequently feed all four methods the same updated guess
the following iteration. This decision precluded a comparison of convergence rate, since if the
methods were allowed to solve a problem at their individual pace, differences in accuracy would
produce differences in guess updates and therefor the number and nature of iterations perfomed.
Different guesses from iterations started with different previous guesses could not be data for a
valid comparison of tangent-stiffness accuracy between methods.  However a separate convergence rate
comparison study could be done given some modifications to the materials for the study described in
this paper, and of course, the resources to perform that study.   Additionally, having each of the
methods physically operate in the same process, single or one of associated parallel, allowed the
comparison of tangent-stiffness matrices as loaded from random access memory rather than from the
hard-disk.  Loading from RAM gives the benefit of vastly greater speed and the simpler programming
compared to some other solution speculatively involving dynamic file management on files which for
one of the components of the study would be on the order of $1E10$ bytes. However, the price of
running the methods together in the same process and avoiding the use of the hard-disk was that at
least two tangent-stiffness matrices had to be stored in RAM during simulation, which meant that
special high-memory compute nodes were needed for the $1E6$ peridynamic element problem series.

@@

Mesh generation for the test problems run in the study was done using the software package
\emph{Cubit} developed at Sandia National Labs \cite{ref-Cubit}. For the test series in which peridynamic node density was
varied, journal scripts compatible with \emph{Cubit} were used. The rule used to select the
arangement of nodes within the borders designated by simulated material block's dimensions was to
manually choose a node spacing, fit the maximum whole number of nodes in a regular grid within the
block, and then check whether or not this number of nodes matched the number of nodes desired for a
particular test. The same process was used for the multicore series. An example journal script can
be found on the corresponding author's website. For future work, it may be beneficial to automate
and randomize this process for time efficiency and to assist statistical analysis.

@@

\subsubsection{Methods: Reduction of Data} The output produced by \emph{Perdigm} including the
additional accuracy and speed measurements it was modified to provide was redirected from the
console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).
Directories and filenames were chosen so that data would be easy to sort by individual test run.
Extraneous information was manually deleted from copied versions of the output text files so that
they would be regular enough to be easily read by scripts written in \emph{Python}. These scripts
averaged accuracy or speed data for a test corresponding to a particular peridynamic node density
over all iterations for that run, and then plotted this averaged data as if it were a function of
peridynamic node density as it varied from test to test in the series.  Similarly, for the multicore
tests, data was averaged and plotted in much the same way, but as a funciton of number of cores used
to solve the problem rather than as a function peridynamic node density which was held constant.
These data reduction and plotting scripts can be found on the corresponding autor's website.

@@

\subsection{Accuracy Data}
Accuracy measurements where taken according to the methods mentioned in \ref{tcs}. The averaged
results of these accuracy measurements for the single core tests can be seen in
\ref{fig:serial_accuracy}. Averaged Frobenius norm of the element-wise difference between a given
Jacobian and the Jacobian produced by the automatic-differentiation method is plotted against number
of nonzero Jacobian matrix elements. Complex-step was revealed as the most accurate of the methods
besides automatic-differentiation, complex-step resulting in several orders of magnitude less
difference than resulted from using the central-difference method and the most inaccurate
forward-difference method. This accuracy ranking order was mantained for each test in the series. It
should be pointed out compared to the L2 norms of any of the Jacobian matrices by themselves, the
magnitude of these differences for any of the methods is relatively small. These values appear in
the experimental data which is available on the corresponding author's website. The effect of these
accuracy differences on convergence rate directly was not determined.

@@

\subsection{Final Thoughts on Complex-Step} Complex-step was shown to be highly accurate in
calculating tangent-stiffness matrices by the standards of this study, but as it is implemented in
\emph{Peridigm} may not be any more accurate than AD, and it is almost certainly slower for a naive
implementation. When looking to the future of \emph{Peridigm}, Complex-step holds the advantage of
relying on byte-copiable datatypes, while AD requires a complicated serialization and
deserialization in order to function as byte-copiable. This is fine for copying in between MPI
processes, but deserialization on an accelerator or GPU is very difficult, as this would require
compiling a version of the AD library being used that were compatible with the accelerator or GPU
being used. What this means is that if \emph{Peridigm} or any simulation software were to be
reformulated to take advantage of accelerators or GPUs, attractive given their performance and power
efficiency advantages over traditional CPUs, the flexibilty of complex-step owed to its simple
implementation may make it more viable for that application.
